Visualisation of Brain Statistics with R-packages _ggseg_ and _ggseg3d_

### 2020-04-06 codethrough.Rmd
add code through

---
title: Visualisation of Brain Statistics with R-packages _ggseg_ and _ggseg3d_
preprint: true
author: 
  - name: Athanasia M. Mowinckel
    affiliation: 1
    corresponding: true
    email: a.m.mowinckel@psykologi.uio.no
  - name: Didac Vidal-Piñeiro
    affiliation: 1
affiliation:
  - code: 1
    address: Center for Lifespan Changes in Brain and Cognition, University of Oslo, PO. box 1094 Blindern, 0317 Oslo, Norway
abstract: >
  There is an increased emphasis on visualizing neuroimaging results in more intuitive ways. Common statistical tools for dissemination, such as bar charts, lack the spatial dimension that is inherent in neuroimaging data. Here we present two packages for the statistical software R, _ggseg_ and _ggseg3d_, that integrate this spatial component. The _ggseg_ and _ggseg3d_ packages visualize pre-defined brain segmentations as both 2D polygons and 3D meshes, respectively. Both packages are integrated with other well-established R-packages, allowing great flexibility. In this tutorial, we present the main data and functions in the _ggseg_ and _ggseg3d_ packages for brain atlas visualization. The main highlighted functions are able to display brain segmentation plots in R. Further, the accompanying _ggsegExtra_-package includes a wider collection of atlases, and is intended for community-based efforts to develop more compatible atlases for _ggseg_ and _ggseg3d_. Overall, the _ggseg_-packages facilitate parcellation-based visualizations in R, improve and ease the dissemination of the results, and increase the efficiency of the workflows. 
bibliography: references.bib
output:
  bookdown::pdf_book:
    base_format: rticles::peerj_article # for using bookdown features like \@ref()
    citation_package: natbib
    fig_caption: true
    includes:
      in_header: preamble.tex
  rticles::peerj_article: preprint
keep_tex: true
---

```{r setup, include=F}
knitr::opts_chunk$set(eval=TRUE, echo=TRUE, 
                      collapse = TRUE, results = 'hold', 
                      cache=FALSE, fig.pos = 'H') 
options(tinytex.verbose = TRUE, tinytex.clean = FALSE,
        tibble.width = 50)

library(dplyr)
library(tidyr)
library(plotly) 
library(ggplot2)
library(ggseg)
library(ggseg3d)
library(ggsegExtra)
library(kableExtra)
library(knitr)

include_plotly <- function(plot, file){
  orca(plot, file = file, format = "png")
  knitr::include_graphics(file)
}

# Hook for setting latex text sizes
def.chunk.hook  <- knitr::knit_hooks$get("chunk") 
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

# Hook for setting code chunk wrapping
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})

get_data_list <- function(package){
  as_tibble(data(package=package)$results)
}

pkglist <- c("ggseg", "ggseg3d",  
             basename(ggsegExtra::ggseg_atlas_repos()$repo))
dt <- lapply(pkglist, get_data_list) %>% 
  bind_rows() %>% 
    filter(!grepl("pal", Item)) %>% 
  mutate(Type = ifelse(grepl("3d", Item), "Mesh", "Polygon"),
         Item = gsub("_3d", "", Item),
         Item = gsub("_", "", Item)) %>% 
    select(Package, Title, Type, Item) %>% 
  spread(Type, Package, fill = "None")  %>% 
  mutate(Citation = case_when(
    grepl("dk", Item) ~ "\\citet{dk}",
    grepl("yeo", Item) ~ "\\citet{yeo2011}",
    grepl("aseg|hcpa", Item) ~ "\\citet{aseg}",
    grepl("desterieux", Item) ~ "\\citet{desterieux}",
    grepl("^chen", Item) ~ "\\citet{chen}",
    grepl("glasser", Item) ~ "\\citet{glasser}",
    grepl("icbm", Item) ~ "\\citet{icbm}",
    grepl("jhu", Item) ~ "\\citet{jhu}",
    grepl("schaefer", Item) ~ "\\citet{schaefer}",
    grepl("tracula", Item) ~ "\\citet{tracula}",
    grepl("^ho", Item) ~ "\\citet{ho}",
    TRUE ~ ""
  ))

options(warnPartialMatchAttr = FALSE,
        warnPartialMatchDollar = FALSE,
        warnPartialMatchArgs = FALSE)
```

@@

The _ggseg_ and _ggseg3d_ packages present compiled data-sets, tailored functions that allow brain data integration and plotting, and other minor features such as custom colour palettes.
The data featured in the packages are derived from two well-known parcellations: the Desikan-Killany cortical atlas (DK; @dk), which covers the cortical surface of the brain, and the Automatic Segmentation of Subcortical Structures (aseg; @aseg), which covers the subcortical structures.
Both atlases are implemented in several neuroimaging softwares, such as FreeSurfer (@fischl_99, @dale_99, @Fischl2000), and are commonly used in relation to developmental changes, disease biomarkers, genomic data, and cognition (@amlien_elaboration_2019, @WALHOVD20051261, @Pizzagalli).
The _ggsegExtra_ package contains a list of online repositories of pre-compiled atlases (currently `r sum(!grepl("None", dt$Polygon))-1` additional polygon atlases and `r sum(!grepl("None", dt$Mesh))-1` mesh atlases) and it is frequently updated.
A summary of all available atlases compatible with the _ggseg_-packages as of `r format(Sys.Date(), "%B %Y")` can be found in Table \@ref(tab:table1).

```{r 'table1', echo=F, warning=FALSE, message=FALSE}
dt %>%
  kable(format="latex", escape = FALSE, booktabs = TRUE,
        caption = "Table of currently available atlases in ggseg and ggseg3d, and listed in the ggsegExtra R-package. Polygon and Mesh refer to 2D and 3D brain atlas representations, respectively. Cell contents in the Mesh and Polygon columns indicate the source package containing the atlas data-set.") %>%
  kable_styling(latex_options = c("scale_down"))
```

@@

```{r "figure7-show", message=FALSE, echo=TRUE, results = 'hide',  out.width='30%',fig.show='hold',fig.ncol = 3}
# remotes::install_github("LCBC-UiO/ggseg3d")
library(ggseg3d)

@@

```{r "figure7", message=FALSE, echo=F, eval=TRUE, out.width='30%',fig.show='hold',fig.ncol = 3, fig.cap='(ref:figure7-cap)'}
orca(fig7l, file = "msc_ggseg_files/figure-latex/figure7l.png", format = "png")
orca(fig7m, file = "msc_ggseg_files/figure-latex/figure7m.png", format = "png")
orca(fig7r, file = "msc_ggseg_files/figure-latex/figure7r.png", format = "png")

include_graphics(c("msc_ggseg_files/figure-latex/figure7l.png",
                   "msc_ggseg_files/figure-latex/figure7m.png",
                   "msc_ggseg_files/figure-latex/figure7r.png"))
```

@@

```{r out.width='80%', linewidth=100, size='small', eval=TRUE}
# Select surface and hemisphere, and then unnest to inspect the atlas data
dk_3d %>%
  filter(surf == "inflated" & hemi == "right") %>%
  unnest(ggseg_3d) %>%
  head(5)
```

@@

```{r 'figure8-show', include=FALSE, out.width='40%', results='hide'}
# Figure 8 left
fig8l <- ggseg3d(.data = someData, 
                 atlas = dk_3d, colour = "p", text = "p") %>% 
  pan_camera("right lateral")
fig8l

@@

```{r 'figure8', out.width='30%', echo=FALSE, fig.cap="(ref:figure8-cap)", fig.show='hold', fig.ncol = 3}
orca(fig8l, file = "msc_ggseg_files/figure-latex/figure8l.png", format = "png")
orca(fig8m, file = "msc_ggseg_files/figure-latex/figure8m.png", format = "png")
orca(fig8r, file = "msc_ggseg_files/figure-latex/figure8r.png", format = "png")

include_graphics(c(
  "msc_ggseg_files/figure-latex/figure8l.png",
  "msc_ggseg_files/figure-latex/figure8m.png",
  "msc_ggseg_files/figure-latex/figure8r.png"))

@@

# Discussion

The main aim of the _ggseg_, _ggseg3d_, and _ggsegExtra_ packages is to ease and streamline visualization of brain atlas data in R, by gathering a collection of atlases from several scientific sources and providing customized plotting functions.
In this tutorial, we introduced the packages to the readers by presenting some use examples and highlighting the main functions and options that are available.
As visualization tools, these packages add up to manifold functionalities such as ggBrain (@ggBrain) and ggneuro (@ggneuro) in R, and software-specific image viewers such as FSLeyes (@fsleyes) and Freeview (@dale_99).
In this regard, we do not aim to compete with software-specific visualizations or advocate for the superiority of the _ggseg_-packages as visualization tools.
After all, flattened 2D polygons do not rely on a meaningful brain coordinate system and the units of information in 3D meshes are limited to the number of parcellations.
On the contrary, we believe the _ggseg_ niche among visualization tools resides in its simplicity and its ability to be combined with statistical analysis pipelines.
The possibility to serve as an interactive tool for dissemination and reproducibility when combined with other technologies, such as Binder (@binder) or Shiny (@shiny), is an added benefit.
This is exemplified in the [online supplementary information](https://athanasiamo.shinyapps.io/Virtual_histology_2019/) of Vidal-Piñeiro et al. (-@vidal_2019).

The functions in the _ggseg_-packages provide users with the possibility of adapting the plots to their wishes, and also makes it possible to create and contribute to the atlas repository in _ggsegExtra_.
The three _ggseg_-packages contain three main features:

1. A collection of 2D-polygon and 3D mesh brain parcellation atlases. The atlas data include the necessary coordinates for plotting, and include other information that should be recognizable for users.  
2. `ggseg()` and `ggseg3d()` functions for visualization. Both functions are flexible and well-adapted to their environment and can be combined with any additional argument from _ggplot2_ and _plotly_, respectively.
`ggseg()` is a wrapper function for `geom_polygon` from _ggplot2_ and it can be built upon like any _ggplot2_ object. 
`ggseg3d()` is a _plotly_ wrapper function for tri-surface mesh plots which prints 3D atlases.  
3. Complimentary features, like color scales, and functions to convert compatible data frames to atlas-classes, read in likely data sources (like Freesurfer analysis output), and help create new atlases for use with _ggseg_ and _ggseg3d_.

The foundations of the _ggseg_-packages trace back to the necessity of visualizing and exploring the lifespan trajectories of cortical thickness across different brain regions (see supplementary information in Vidal-Piñeiro et al. (-@vidal_2019) ) .
That is, _ggseg_ appears with the need to inspect and display brain information over time - i.e. including a spatial dimension and a time-varying factor - overcoming the constrains of printed journals and classical 2D plots (e.g. bar plots).
The current state of science requires researches to share the results of studies in both high detail and in an intuitive manner, as it permits communication to wide audiences and facilitates reproducibility.
Hence, we believe this tool conforms to the essence of open science and invite users to improve the code, provide examples, or tutorials, and contribute to the atlas collection according to their own interest and needs via the public [ggseg GitHub repository](https://github.com/LCBC-UiO/ggseg), [ggseg3d GitHub repository](https://github.com/LCBC-UiO/ggseg3d), and [ggsegExtra GitHub repository](https://github.com/LCBC-UiO/ggsegExtra).

The ggseg-packages are based on _ggplot2_ and _plotly_, which are only two of several visualisation tools in R, in addition to base-R plotting. 
While the functionality presented here could also be integrated into packages such as [lattice](https://www.rdocumentation.org/packages/lattice/versions/0.20-40), both _ggplot2_ and _plotly_ have become established tools in the R-community and have extensive online documentation.
This makes them great packages to base further extensions and wrappers on.

Finally - while the _ggseg_-packages are circumscribed to brain parcellations - we believe that the structure and functions of the package can be easily applied to any scientific field that benefits from data being displayed across the spatial dimension.
We encourage readers to borrow the package functionalities and adapt it to their respective fields and structures of interest, such as has already been done with the _gganatogram_-package (@gganatogram).

# Planned package improvements
In terms of planned _ggseg_ improvements, we are working on making a proper `geom` or `stat`.
While `ggseg()` works quite nicely as a `ggplot` wrapper, it introduces some unexpected work for the user, like having to group data _before_ faceting. 
A `geom` or `stat` would benefit the user as it would behave more like other _ggplot2_ `geoms` and there would be less new functionalities to learn.

Concerning _ggseg3d_, we also hope to make it possible to colour each individual mesh triangle, and as such provide the users with more flexibility in the plotting. 
This should also improve the speed of `ggseg3d()`, which currently is quite slow due to each parcellation region being loaded into `plotly` sequentially. 
Loading the entire brain mesh at once, and then coloring mesh triangles would significiantly improve the speed of `ggseg3d()`. 
When able to colour the mesh triangles directly, rather than each parcellation individually, only full brain representations of each surface (rather than each surface of each atlas) needs to be stored, in addition to text data indexing mesh-locations.
This would reduce package size considerably, including package sizes of all atlas-repositories.

# Conclusion
Visualization is a fundamental aspect of neuroimaging to explore and understand data, guide interpretation, and communicate with colleagues and the general audience.
In this tutorial, we have introduced the _ggseg_-packages, tools for visualizing brain statistics through brain parcellation atlases in R.
This visualization tool easily combines with interactive routines as well as with diverse statistical analysis pipelines.
We hope this tool and tutorial proves useful to neuroscientists and inspires others to apply the functions in a wide variety of fields and structures.

# Author Contributions
Didac Vidal-Piñeiro generated the idea for the tool, and the initial scripts for plot visualization.
He has also been responsible for converting images from neuroimaging data to ggseg-like data (e.g. polygons and mesh data).
Athanasia M. Mowinckel adapted the initial scripts and made the functions into package format and has continued developing the functions with the aim of increasing user-friendliness.
She is also responsible for conceiving and adding the mesh-plot functionality through plotly, and developing the pipeline for making that possible.
A. M. Mowinckel wrote the first draft of the paper, and both have since critically edited it.

# Conflicts of Interest
The authors declare that there were no conflicts of interest with respect to the authorship or the publication of this article.

# Acknowledgements
We thank John Muschelli for package code comments and its adaptation to neuroconductor ([2018]), and Richard Beare for the first community created atlas, and base code comments.
We are indebted to A. M. Winkler scripts for converting neuroimaging data (Freesurfer) into .ply files, a necessary step for neuroimaging to mesh-plots conversion, and to Inge Amlien for the ‘LCBC’ surface files and helpful contributions to the ggseg3d pipeline.
We are indebted to those users that actively contributed to the package and all the individuals that worked on the framework on which the package is sustained (e.g. R-project, neuroimaging software).
Finally, we thank Anders M. Fjell (AMF) and Kristine B. Walhovd (KBW) for their encouragement and financial support.

>>>>

---
title: "Visualisation of Brain Statistics with R-packages _ggseg_ and _ggseg3d_"
subtite: "code through"
output: html_notebook
---

```{r setup, include=F}
knitr::opts_chunk$set(eval=TRUE, echo=TRUE, 
                      collapse = TRUE, results = 'hold', 
                      cache=FALSE, fig.pos = 'H') 
options(tinytex.verbose = TRUE, tinytex.clean = FALSE,
        tibble.width = 50) 
```

@@

The _ggseg_ and _ggseg3d_ packages present compiled data-sets, tailored functions that allow brain data integration and plotting, and other minor features such as custom colour palettes.
The data featured in the packages are derived from two well-known parcellations: the Desikan-Killany cortical atlas (DK; @dk), which covers the cortical surface of the brain, and the Automatic Segmentation of Subcortical Structures (aseg; @aseg), which covers the subcortical structures.
Both atlases are implemented in several neuroimaging softwares, such as FreeSurfer (@fischl_99, @dale_99, @Fischl2000), and are commonly used in relation to developmental changes, disease biomarkers, genomic data, and cognition (@amlien_elaboration_2019, @WALHOVD20051261, @Pizzagalli).
The _ggsegExtra_ package contains a list of online repositories of pre-compiled atlases (currently `r sum(!grepl("None", dt$Polygon))-1` additional polygon atlases and `r sum(!grepl("None", dt$Mesh))-1` mesh atlases) and it is frequently updated.

@@

```{r "figure7-show", out.width='30%',fig.show='hold',fig.ncol = 3}
# remotes::install_github("LCBC-UiO/ggseg3d")
library(ggseg3d)

@@

@@

```{r out.width='80%'}
# Select surface and hemisphere, and then unnest to inspect the atlas data
dk_3d %>%
  filter(surf == "inflated" & hemi == "right") %>%
  unnest(ggseg_3d) %>%
  head(5)
```

@@

```{r 'figure8-show', include=FALSE, out.width='40%'}
# Figure 8 left
fig8l <- ggseg3d(.data = someData, 
                 atlas = dk_3d, colour = "p", text = "p") %>% 
  pan_camera("right lateral")
fig8l

@@

@@
